---
title: "Validation of a scale"
author: "Kushan De Silva"
date: "June 23, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(psych)
data(sat.act)
describe(sat.act) #basic descriptive statistics
```


```{r}
pairs.panels(sat.act)
lowerCor(sat.act)
cor(sat.act)

female <- subset(sat.act,sat.act$gender==2)
male <- subset(sat.act,sat.act$gender==1)
lower <- lowerCor(male[-1])

pairs.panels(sat.act,pch='.')
upper <- lowerCor(female[-1])
both <- lowerUpper(lower,upper)
round(both,2)

pairs.panels(sat.act,pch='.')

diffs <- lowerUpper(lower,upper,diff=TRUE)
round(diffs,2)

cor.plot(Thurstone,numbers=TRUE,main="9 cognitive variables from Thurstone")

circ <- sim.circ(24)
r.circ <- cor(circ)
cor.plot(r.circ,main='24 variables in a circumplex')

```


In FA, the goal is to identify fewer underlying constructs to explain the observed data. In PCA, the goal can be mere data reduction, but the interpretation of components is frequently done in terms similar to those used when describing the latent variables estimated by FA. Cluster analytic techniques, although usually used to partition the subject space rather than the variable space, can also be used to group variables to reduce the complexity of the data by forming fewer and more homogeneous sets of tests or items.


Minimum Residual Factor Analysis, Principal Axis Factor Analysis, Weighted Least Squares Factor Analysis

```{r}
f3t <- fa(Thurstone,3,n.obs=213)
f3t

f3 <- fa(Thurstone,3,n.obs = 213,fm="pa")
f3o <- target.rot(f3)
f3o

f3w <- fa(Thurstone,3,n.obs = 213,fm="wls")
print(f3w,cut=0,digits=3)

plot(f3t)

fa.diagram(f3t)
fa.diagram(f3w)
fa.diagram(f3o)


```


Although the goals of PCA and FA are similar, PCA is a descriptive model of the data, while FA is a structural model.PCA model attempts to account for the entire variance of the correlation matrix, while FA accounts for just the common variance.

```{r}
p3p <-principal(Thurstone,3,n.obs = 213,rotate="Promax")
p3p
```

```{r}
om.h <- omega(Thurstone,n.obs=213,sl=FALSE)
op <- par(mfrow=c(1,1))

om <- omega(Thurstone,n.obs=213)


```



Item cluster Analysis

the iclust algorithm is particularly appropriate for scale construction.The iclust forms clusters of items using a hierarchical clustering algorithm until one of two measures of internal consistency fails to increase.


```{r}
data(bfi)
ic <- iclust(bfi[1:25])

summary(ic) #show the results

data(bfi)
r.poly <- polychoric(bfi[1:25])

f4 <- fa(bfi[1:25],4,fm="pa")
factor.congruence(f4,ic)

ic.poly <- iclust(r.poly$rho,title="ICLUST using polychoric correlations")
iclust.diagram(ic.poly)

ic.poly <- iclust(r.poly$rho,5,title="ICLUST using polychoric correlations for nclusters=5")
iclust.diagram(ic.poly)

ic.poly <- iclust(r.poly$rho,beta.size=3,title="ICLUST beta.size=3")
iclust.diagram(ic.poly)

print(ic,cut=.3)

fa(bfi[1:10],2,n.iter=20)

factor.congruence(list(f3t,f3o,om,p3p))

vss <- vss(bfi[1:25],title="Very Simple Structure of a Big 5 inventory")
vss
plot(vss)

fa.pa <- fa.parallel(bfi[1:25],main="Parallel Analysis of a Big 5 inventory")

v16 <- sim.item(16)
s <- c(1,3,5,7,9,11,13,15)
f2 <- fa(v16[,s],2)
fe <- fa.extension(cor(v16)[s,-s],f2)
fa.diagram(f2,fe=fe)


```



Reliability of a single scale

```{r}
set.seed(17)
r9 <- sim.hierarchical(n=500,raw=TRUE)$observed
round(cor(r9),2)

alpha(r9)

keys <- c(1,-1,1,1,1,1,1)
alpha(attitude,keys)

keys <- c(1,1,1,1,1,1,1)
alpha(attitude,keys)

items <- sim.congeneric(N=500,short=FALSE,low=-2,high=2,categorical=TRUE) #500 responses to 4 discrete items
alpha(items$observed) #item response analysis of congeneric measures


```



Using omega to nd the reliability of a single scale

```{r}
om.9 <- omega(r9,title="9 simulated variables")
om.9


```

Estimating omega using Confirmatory Factor Analysis

The omegaSem function will do an exploratory analysis and then take the highest loading
items on each factor and do a conrmatory factor analysis using the sem package.


```{r}
om9 <- omegaSem(r9,n.obs=500)
om9

```


Other estimates of reliability

```{r}
splitHalf(r9)

```

```{r}
keys <- make.keys(nvars=28,list(Agree=c(-1,2:5),Conscientious=c(6:8,-9,-10), Extraversion=c(-11,-12,13:15),Neuroticism=c(16:20), 
                                Openness = c(21,-22,23,24,-25)),
                  item.labels=colnames(bfi))
keys


keys.1<- make.keys(10,list(Agree=c(-1,2:5),Conscientious=c(6:8,-9,-10)))
keys.2 <- make.keys(15,list(Extraversion=c(-1,-2,3:5),Neuroticism=c(6:10), 
                            Openness = c(11,-12,13,14,-15)))
keys.25 <- superMatrix(list(keys.1,keys.2))

scores <- scoreItems(keys,bfi)
scores

```


Forming scales from a correlation matrix

```{r}
r.bfi <- cor(bfi,use="pairwise")
scales <- cluster.cor(keys,r.bfi)
summary(scales)

pairs.panels(scores$scores,pch='.',jiggle=TRUE)


```

Scoring Multiple Choice Items

```{r}
data(iqitems)
iq.keys <- c(4,4,4, 6,6,3,4,4, 5,2,2,4, 3,2,6,7)
score.multiple.choice(iq.keys,iqitems)

#just convert the items to true or false
iq.tf <- score.multiple.choice(iq.keys,iqitems,score=FALSE)
describe(iq.tf) #compare to previous results


```


Item analysis

```{r}
data(iqitems)
iq.keys <- c(4,4,4, 6,6,3,4,4, 5,2,2,4, 3,2,6,7)
scores <- score.multiple.choice(iq.keys,iqitems,score=TRUE,short=FALSE)
op <- par(mfrow=c(2,2)) #set this to see the output for multiple items
irt.responses(scores$scores,iqitems[1:4],breaks=11)


```

Factor analysis and Item Response Theory

```{r}
set.seed(17)
d9 <- sim.irt(9,1000,-2.5,2.5,mod="normal") #dichotomous items
test <- irt.fa(d9$items)
test

data(bfi)
e.irt <- irt.fa(bfi[11:15])
e.irt


op <- par(mfrow=c(3,1))
plot(test,type="ICC")
plot(test,type="IIC")
plot(test,type="test")
op <- par(mfrow=c(1,1))

e.info <- plot(e.irt,type="IIC")
print(e.info,sort=TRUE)

iq.irt <- irt.fa(iq.tf)
iq.irt

v9 <- sim.irt(9,1000,-2.,2.,mod="normal") #dichotomous items
items <- v9$items
test <- irt.fa(items)
om <- omega(iq.irt$rho,4)
om

total <- rowSums(items)
ord <- order(total)
items <- items[ord,]
#Delete some of the data - note that they are ordered by score
items[1:333,5:9] <- NA
items[334:666,3:7] <- NA
items[667:1000,1:4] <- NA
scores <- score.irt(test,items)
unitweighted <- score.irt(items=items,keys=rep(1,9))
scores.df <- data.frame(true=v9$theta[ord],scores,unitweighted)
colnames(scores.df) <- c("True theta","irt theta","total","fit","rasch","total","fit")


```


Multilevel modeling

```{r}
pairs.panels(scores.df,pch='.',gap=0)
title('Comparing true theta for IRT, Rasch and classically based scoring')

C <- cov(sat.act,use="pairwise")
model1 <- lm(ACT~ gender + education + age, data=sat.act)
summary(model1)


#compare with mat.regress
set.cor(c(4:6),c(1:3),C, n.obs=700)

```

Graphical structure output


```{r}
f3 <- fa(Thurstone,3)
plot(f3)
fa.diagram(f3)
c <- iclust(Thurstone)
plot(c) #a pretty boring plot
iclust.diagram(c) #a better diagram
c3 <- iclust(Thurstone,3)
plot(c3) #a more interesting plot
data(bfi)
e.irt <- irt.fa(bfi[11:15])
plot(e.irt)
ot <- omega(Thurstone)
plot(ot)
omega.diagram(ot)
```

```{r}
xlim=c(0,10)
ylim=c(0,10)
plot(NA,xlim=xlim,ylim=ylim,main="Demontration of dia functions",axes=FALSE,xlab="",ylab="")
ul <- dia.rect(1,9,labels="upper left",xlim=xlim,ylim=ylim)
ll <- dia.rect(1,3,labels="lower left",xlim=xlim,ylim=ylim)
lr <- dia.ellipse(9,3,"lower right",xlim=xlim,ylim=ylim)
ur <- dia.ellipse(9,9,"upper right",xlim=xlim,ylim=ylim)
ml <- dia.ellipse(3,6,"middle left",xlim=xlim,ylim=ylim)
mr <- dia.ellipse(7,6,"middle right",xlim=xlim,ylim=ylim)
bl <- dia.ellipse(1,1,"bottom left",xlim=xlim,ylim=ylim)
br <- dia.rect(9,1,"bottom right",xlim=xlim,ylim=ylim)
dia.arrow(from=lr,to=ul,labels="right to left")
dia.arrow(from=ul,to=ur,labels="left to right")
dia.curved.arrow(from=lr,to=ll$right,labels ="right to left")
dia.curved.arrow(to=ur,from=ul$right,labels ="left to right")
dia.curve(ll$top,ul$bottom,"double") #for rectangles, specify where to point
dia.curved.arrow(mr,ur,"up") #but for ellipses, just point to it.
dia.curve(ml,mr,"across")
dia.arrow(ur,lr,"top down")
dia.curved.arrow(br$top,lr$bottom,"up")
dia.curved.arrow(bl,br,"left to right")
dia.arrow(bl,ll$bottom)
dia.curved.arrow(ml,ll$right)
dia.curved.arrow(mr,lr$top)


```

